Structural basis for molecular assembly of fucoxanthin chlorophyll a/c-binding proteins in a diatom photosystem I supercomplex

Version 1

Abstract

Abstract

Photosynthetic organisms display considerable diversity in light-harvesting complexes (LHCs). LHCs are attached to photosystem I (PSI), contributing to the formation of the PSI-LHCI supercomplex. The number of LHCIs and their protein and pigment compositions have been found to differ greatly among the PSI-LHCI structures. However, it remains unclear how LHCIs recognize their specific binding sites in the PSI core. In this study, we elucidated the cryo-electron microscopic structure of a PSI supercomplex incorporating fucoxanthin chlorophyll a/c-binding proteins (FCPs), designated as PSI-FCPI, isolated from the diatom Thalassiosira pseudonana CCMP1335. The structural analysis of PSI-FCPI revealed a composition of five FCPI subunits associated with a PSI monomer, specifically identified as RedCAP, Lhcr3, Lhcq10, Lhcf10, and Lhcq8. Through structural and sequence analyses, we identified distinct protein-protein interactions at the interfaces between FCPI and PSI subunits, as well as among FCPI subunits themselves. Comparative structural analyses of PSI-FCPI supercomplexes and phylogenetic analysis of FCPs across T. pseudonana and the diatom Chaetoceros gracilis highlight the evolutionary conservation of protein motifs crucial for the selective binding of individual FCPI subunits. These findings significantly advance our understanding of the molecular mechanisms governing the assembly and selective binding of FCPIs in diatoms.

Introduction

Oxygenic photosynthesis in cyanobacteria, algae, and land plants converts solar energy into chemical energy concomitant with the evolution of oxygen molecules (1). The light-energy conversion reactions occur within two multi-subunit membrane protein complexes known as photosystem I and photosystem II (PSI and PSII, respectively), which perform light harvesting, charge separation, and electron transfer reactions (2–5). To harvest light energy efficiently, a number of light-harvesting antenna subunits are attached to the periphery of PSI and PSII core complexes and transfer excitation energy to the two types of photosystem cores (1). Light-harvesting antennae have a wide variety among photosynthetic organisms in relation to protein sequences and pigment compositions and can be categorized into two major groups: membrane proteins and water-soluble proteins (1).

The membrane protein category is composed mainly by the light-harvesting complex (LHC) protein superfamily (6, 7), which absorbs light energy using chlorophylls (Chls) and carotenoids (Cars). The number and types of Chls and Cars are highly diversified among LHCs, resulting in color variations in photosynthetic organisms, which can be classified into green and red lineages (8). The green lineage organisms include green algae and land plants, whereas the red lineage organisms include red algae, diatoms, haptophytes, cryptophytes, and dinoflagellates (8). LHCs that are specific to PSI (LHCIs) interact with a eukaryotic PSI monomer, thereby forming a PSI-LHCI supercomplex (9, 10), whose structures have been revealed by cryo-electron microscopy (cryo-EM) in a number of eukaryotes (9, 10). In the red lineage, the number of LHCIs and their protein and pigment compositions have been found to differ greatly among the PSI-LHCI structures of red algae (11–14), a diatom (15, 16), a cryptophyte (17), and dinoflagellates (18, 19).

We have recently shown the conservation and diversity of LHCIs among red-lineage algae based on structural and phylogenetic analyses of PSI-LHCI supercomplexes (14). This work clearly showed that the binding sites of LHCIs to PSI were conserved to some extent among red-lineage algae but their evolutionary relationship was poor. It is known that LHCIs have similar protein structures, especially their three-transmembrane helices, among photosynthetic organisms irrespective of the green and red lineages (9, 10); nevertheless, individual LHCIs have changed their sequences and structures to adapt their binding sites to the PSI cores during assembly of the PSI-LHCI supercomplexes. These observations raise a critical question as to how LHCIs recognize their binding sites in the PSI core.

Diatoms are one of the most important phytoplankton in aquatic environments and contribute to the primary production in the ocean remarkably (20). Diatoms have unique LHCs called fucoxanthin Chl a/c-binding proteins (FCPs), whose pigment compositions and amino acid sequences are different from those of LHCs in land plants (21–23). Isolation and characterization of PSI-FCPI supercomplexes isolated from the diatom Chaetoceros gracilis and their structures have been reported (15, 16, 24–29). Kumazawa et al. showed the high diversity of FCPs between two types of diatoms, C. gracilis and Thalassiosira pseudonana (30). C. gracilis and T. pseudonana have 46 and 44 FCPs, respectively, which were categorized into multiple, closely related subgroups (30), and the amino acid sequences of FCPs are not completely identical between the two diatoms. Therefore, comparing FCPIs including their amino acid residues and protein structures at similar binding sites in PSI-FCPIs between the two diatoms may lead to molecular insights into how FCPIs are bound to PSI. However, an overall structure of the T. pseudonana PSI-FCPI has not been solved yet.

In this study, we solved a PSI-FCPI structure of T. pseudonana CCMP1335 at a resolution of 2.30 Å by cryo-EM single-particle analysis. The structure shows a PSI-monomer core and five FCPI subunits. Structural and sequence comparisons exhibit unique protein-protein interactions of each FCPI subunit with PSI. Based on these findings, we discuss the molecular assembly and selective binding mechanisms of FCPI subunits in diatom species.

Results and discussion

Overall structure of the T. pseudonana PSI-FCPI supercomplex

The PSI-FCPI supercomplexes were purified from the diatom T. pseudonana CCMP1335 as described in the Methods section. Biochemical and spectroscopic analyses showed that this complex is intact with several bands corresponding to FCPI (Fig. S1). Cryo-EM images of the PSI-FCPI supercomplex were obtained by a JEOL CRYO ARM 300 electron microscope operated at 300 kV. The final cryo-EM map was determined at a resolution of 2.30 Å with a C1 symmetry (Fig. S2, S3, Table S1), based on the “gold standard” Fourier shell correlation (FSC) = 0.143 criterion (Fig. S3A).

The atomic model of PSI-FCPI was built based on the resultant cryo-EM map (see Methods; Fig. S3, Table S1–S3). The structure reveals a monomeric PSI core associated with five FCPI subunits (Fig. 1A, B). The five FCPI subunits were named FCPI-1 to 5 (Fig. 1A) following the naming of LHCI subunits in the PSI-LHCI structure of Cyanidium caldarium RK-1 (NIES-2137) (14); in particular, the sites of FCPI-1 and FCPI-2 in the T. pseudonana PSI-FCPI structure (Fig. 1A) corresponded to those of LHCI-1 and LHCI-2 in the C. caldarium PSI-LHCI structure. The PSI core contains 94 Chls a, 18 β-carotenes (BCRs), 1 zeaxanthin (ZXT), 3 [4Fe-4S] clusters, 2 phylloquinones, and 6 lipid molecules, whereas the five FCPI subunits contain 45 Chls a, 7 Chls c, 2 BCRs, 15 fucoxanthins (Fxs), 7 diadinoxanthins (Ddxs), and 3 lipid molecules (Table S3).

Overall structure of the PSI-FCPI supercomplex from T. pseudonana.

Structures are viewed from the stromal side (left panels) and the direction perpendicular to the membrane normal (right panels). Only protein structures are shown, and cofactors are omitted for clarity. The FCPI (A) and PSI-core (B) subunits are labeled and colored differently. The five FCPI subunits are labeled as FCPI-1 to 5 (red) with their gene products indicated in parentheses (black) in panel A.

Structure of the T. pseudonana PSI core

The PSI core contains 12 subunits, 11 of which are identified as PsaA, PsaB, PsaC, PsaD, PsaE, PsaF, PsaI, PsaJ, PsaL, PsaM, and Psa29 (Fig. 1B), whereas the remaining one could not be assigned and was modeled as polyalanines because of insufficient map resolution for identification of this protein (Fig. S4A). The unidentified subunit was named Unknown, which is located at the same site as Psa28 in the C. gracilis PSI-FCPI (15). The protein structure of Unknown is virtually identical to that of Psa28 in the C. gracilis PSI-FCPI (Fig. S4B). Psa28 is a novel subunit found in the C. gracilis PSI-FCPI structure (15), and its name follows the nomenclature as suggested previously (31). It is known that the genes encoding various PSI proteins have been designated as psaA, psaB, etc. PsaZ has been identified in PSI of Gloeobacter violaceus PCC 7421 (32, 33). After psaZ, the newly identified genes should be named psa27, psa28, etc., and the corresponding proteins are called Psa27, Psa28, etc. Psa27 has been identified in PSI of Acaryochloris marina MBIC11017 (34–36). Thus, we named the novel subunit as Psa28 (15). In contrast, Psa28 was also named PsaR in the PSI-FCPI structure of C. gracilis (16) and the structure of a PSI supercomplex containing alloxanthin Chl a/c-binding protein (PSI-APCI) from Chroomonas placoidea (17).

Psa29 is newly identified in the T. pseudonana PSI-FCPI structure using ModelAngelo (37) and the NCBI database (https://www.ncbi.nlm.nih.gov/) (Fig. 2). The subunit corresponding to Psa29 was found in the C. gracilis PSI-FCPI structures (15, 16), which was modeled as polyalanines and named as either Unknown1 (15) or PsaS (16), respectively. Psa29 shows a unique structure distinct from the other PSI subunits in the T. pseudonana PSI-FCPI (Fig. 2A), and engages in multiple interactions with PsaB, PsaC, PsaD, and PsaL at distances of 2.5–3.2 Å (Fig. 2B–G). Based on sequence analyses, Psa29 can exhibit evolutionary divergence between Bacillariophyceae (diatoms) and Bolidophyceae, the latter of which is a sister group of diatoms within Stramenopiles (Fig. 2H); however, it was not found from other organisms. The arrangement of each PSI subunit in the T. pseudonana PSI-FCPI is virtually identical to that in the C. gracilis PSI-FCPI structures already reported (15, 16).

Structure and diversity of Psa29.

(A) Structure of Psa29 depicted as cartoons. Psa29 was modeled from V47 to L178. (B) Cryo-EM map of Psa29 and its surrounded environments viewed from the stromal side. The red squared areas are enlarged in panels C–G. Yellow, PsaB; cyan, PsaC; blue, PsaD; magenta, PsaL; orange, Psa29. (C–G) Protein-protein interactions of Psa29 with PsaB/PsaC (C), PsaC/PsaD (D), PsaB/PsaD (E), PsaB/PsaD/PsaL (F), and PsaD (G). Interactions are indicated by dashed lines, and the numbers are distances in Å. Amino acid residues participating in the interactions are labeled; for example, A71/C means Ala71 of PsaC. B, PsaB; C, PsaC; D, PsaD, L, PsaL; 29, Psa29. (H) Phylogenetic analysis of Psa29 in photosynthetic organisms. A maximum-likelihood tree of the Psa29 proteins was inferred by IQ-TREE v2.2.2.7 using WAG+F+G4 model and trimmed alignment of 22 sequences with 245 amino-acid residues. Numbers at nodes are ultrafast bootstrap support (%) (1000 replicates). The tree was mid-point rooted between diatoms and Bolidophyceae Parmales. Psa29 of T. pseudonana CCMP1335 is indicated by a red line.

The number and arrangement of Chls and Cars within the PSI core in the T. pseudonana PSI-FCPI structure (Fig. S4C, D) are largely similar to those in the C. gracilis PSI-FCPI structure (15). However, Chl a102 of PsaI was found in the T. pseudonana PSI-FCPI structure but not in the C. gracilis PSI-FCPI structure (15), whereas a845 of PsaA and BCR863 of PsaB were identified in the C. gracilis PSI-FCPI structure (15) but not in the T. pseudonana PSI-FCPI structure. One of the Car molecules in PsaJ was identified as ZXT103 in the T. pseudonana PSI-FCPI structure but it is BCR112 in the C. gracilis PSI-FCPI structure (15).

Structure of the T. pseudonana FCPIs

Kumazawa et al. classified 44 Lhc genes in T. pseudonana with their characteristic names of Lhcf, Lhcq, Lhcr, Lhcx, and Lhcz (30). Based on this report, the five FCPI subunits in the PSI-FCPI structure were identified using five genes, namely, RedCAP, Lhcr3, Lhcq10, Lhcf10, and Lhcq8, at the sites of FCPI-1 to 5, respectively (Fig. 1A). It should be noted that RedCAP is not included in the 44 Lhc genes (30) but grouped into the LHC protein superfamily (6, 7). For the assignments of each FCPI subunit, we focused on characteristic amino acid residues based on their cryo-EM map, especially S61/V62/Q63 in FCPI-1; A70/R71/W72 in FCPI-2; Y64/R65/E66 in FCPI-3; M63/R64/Y65 in FCPI-4; and A62/R63/R64 in FCPI-5 (Fig. S5). The root mean square deviations (RMSDs) of the structures between FCPI-4 and the other four FCPIs range from 1.91 to 3.73 Å (Table S4).

Each FCPI subunit binds several Chl and Car molecules, namely, 7 Chls a/1 Chl c/2 Fxs/3 Ddxs/2 BCRs in FCPI-1; 10 Chls a/1 Chl c/3 Fxs/1 Ddx in FCPI-2; 7 Chls a/3 Chls c/2 Fxs/2 Ddxs in FCPI-3; 11 Chls a/2 Chls c/4 Fxs in FCPI-4; and 10 Chls a/4 Fxs/1 Ddx in FCPI-5 (Fig. S6, Table S3). The axial ligands of the central Mg atoms of Chls within each FCPI are provided mainly by the main and side chains of amino acid residues (Table S5). Possible excitation-energy-transfer pathways can be proposed based on close physical interactions among Chls between FCPI-3 and PsaA, between FCPI-3 and PsaL, between FCPI-1 and PsaI, and between FCPI-2 and PsaB (Fig. S7).

Structural characteristics of RedCAP and its evolutionary implications

Among the FCPI subunits, only FCPI-1 has BCRs in addition to Fxs and Ddxs (Fig. S6A). This is a first report for the binding of BCRs to FCPIs in diatoms. FCPI-1 is a RedCAP, which belongs to the LHC protein superfamily but is distinct from the LHC protein family (6, 7). FCPI-1 is located near PsaB, PsaI, and PsaL through protein-protein interactions with these subunits at the stromal and lumenal sides (Fig. 3A). I138 and S139 of FCPI-1 interact with K121, G122, and F125 of PsaL at the stromal side (Fig. 3B), whereas at the lumenal side, multiple interactions are found between I109 of FCPI-1 and F5 of PsaI, between T105/L106/T108 of FCPI-1 and W92/P94/F96 of PsaB, and between E102/W103 of FCPI-1 and S71/I73 of PsaL (Fig. 3C). The protein-protein interactions at the lumenal side (Fig. 3C) appear to be caused by a loop structure of FCPI-1 from Q96 to T116 (pink in Fig. 3D), which is characteristic of FCPI-1 but not present in other four FCPI subunits (pink in Fig. 3E). This loop structure is inserted into a cavity formed by PsaB, PsaI, and PsaL (Fig. 3C, D). These findings indicate that the Q96–T116 loop recognizes the cavity for its binding at the specific FCPI-1 site.

Structural characteristics of FCPI-1 (RedCAP).

(A) Interactions of FCPI-1 with PsaB, PsaI, and PsaL viewed from the stromal (left) and lumenal (right) sides. The areas encircled by black squares are enlarged in panels B, C. Yellow, PsaB; magenta, PsaI; dark red, PsaL; red, FCPI-1. (B, C) Protein-protein interactions of FCPI-1 with PsaL (B) and PsaB/PsaI/PsaL (C), respectively. Interactions are indicated by dashed lines, and the numbers are distances in Å. Amino acid residues participating in the interactions are labeled; for example, S139/1 means Ser139 of FCPI-1. B, PsaB; I, PsaI; L, PsaL; 1, FCPI-1. (D) Characteristic loop structure from Q96 to T116 in FCPI-1 viewed from the lumenal side. Q96 and T116 are labeled with sticks and the Q96–T116 loop is colored pink. (E) Superpositions of FCPI-1 (red) with FCPI-2 (green), FCPI-3 (blue), FCPI-4 (magenta), and FCPI-5 (orange). Only proteins are depicted. The Q96–T116 loop of FCPI-1 is colored pink and Q96 and T116 are labeled with sticks.

RedCAP of C. gracilis (CgRedCAP) was not identified in the C. gracilis PSI-FCPI structures (15, 16). As discussed previously (14), we identified CgRedCAP by sequence analysis and suggested the binding of CgRedCAP to the C. gracilis PSI core at the site similar to LHCI-1 of the red alga C. caldarium PSI-LHCI, whose site corresponds to the FCPI-1 site in the PSI-FCPI of T. pseudonana in this study. A sequence alignment between RedCAP of T. pseudonana (TpRedCAP) and CgRedCAP is shown in Fig. S8A, which shows that CgRedCAP has a sequence similarity of 72% to TpRedCAP and has a protein motif of Q106–I113 (QWGTLATI) corresponding to E102–I109 (EWGTLATI) of TpRedCAP (Fig. 3C). This suggests the possible binding of CgRedCAP to PSI of C. gracilis at the position similar to FCPI-1 in the T. pseudonana PSI-FCPI structure. However, it is unknown (i) whether CgRedCAP is indeed bound to the C. gracilis PSI-FCPI supercomplex and (ii) if a loop structure corresponding the Q96–T116 loop of TpRedCAP exists in CgRedCAP. Further structural study of the C. gracilis PSI-FCPI will pave the way for elucidating the mechanism of molecular assembly of diatom RedCAPs.

RedCAPs have been found in the structures of PSI-LHCI of the red alga Porphyridium purpureum (13) and a PSI supercomplex with alloxanthin Chl a/c-binding proteins (PSI-ACPI) of the cryptophyte Chroomonas placoidea (17), which have been summarized in our previous study (14). Both of the RedCAPs of P. purpureum (PpRedCAP) and C. placoidea (CpRedCAP) display loop structures similar to the Q96–T116 loop in TpRedCAP found in the present study (Fig. 8B). Multiple sequence alignments of TpRedCAP with PpRedCAP and CpRedCAP are shown in Fig. 8C. PpRedCAP and CpRedCAP show sequence similarities of 39% and 60%, respectively, to TpRedCAP. PpRedCAP has a protein motif of V105–L112 (VWGPLAQL), while CpRedCAP has a protein motif of Q117–A124 (QWGPLASA). These two motifs correspond to E102–I109 (EWGTLATI) of TpRedCAP; however, the conservation of sequences is lower between TpRedCAP and PpRedCAP/CpRedCAP than between TpRedCAP and CgRedCAP. Among the four RedCAPs, the four amino acids of Trp, Gly, Leu, and Ala are conserved in the protein motifs (xWGxLAxx), implying that the characteristic loop structure including the conserved protein motifs (xWGxLAxx) contributes to the binding of RedCAP to PSI among the red-lineage algae.

Protein-protein interactions of other FCPI subunits

FCPI-2 (Lhcr3) is located near PsaB and PsaM through protein-protein interactions with these subunits at distances of 3.0–4.3 Å at the stromal and lumenal sides (Fig. 4). The amino acid residues of I63/T65/D66/Y69/W134/Y138/D140 in FCPI-2 are associated with W153/L154/K159/F160/W166 in PsaB at the stromal side (Fig. 4B), whereas F116 and F120 of FCPI-2 interact with F5/I9/M12 of PsaM at the lumenal side (Fig. 4C). The amino acid sequences of I63–Y69, F116–F120, and W134–D140 in Lhcr3 are not conserved in the Lhcr subfamily comprising of Lhcr1, Lhcr4, Lhcr7, Lhcr11, Lhcr12, Lhcr14, Lhcr17, Lhcr18, Lhcr19, and Lhcr20 according to Kumazawa et al. (30) (Fig. S9).

Structural characteristics of FCPI-2.

(A) Interactions of FCPI-2 with PsaB and PsaM viewed from the stromal (left) and lumenal (right) sides. The areas encircled by black squares are enlarged in panels B, C. The PSI and FCPI subunits are colored grey and yellow, respectively. Protein-protein interactions are shown as different colors: green, FCPI-2; cyan, PsaB; pink, PsaM. (B, C) Protein-protein interactions of FCPI-2 with PsaB (B) and PsaM (C), respectively. Interactions are indicated by dashed lines, and the numbers are distances in Å. Amino acid residues participating in the interactions are labeled; for example, Y138/2 means Tyr138 of FCPI-2. B, PsaB; M, PsaM.

FCPI-3 (Lhcq10) is located near PsaL through protein-protein interactions with it at distances of 2.3–4.2 Å at the stromal side (Fig. 5A, B). The amino acid residues of L126/I130/L142/Y146/W147/V148/W155 in FCPI-3 are associated with F4/K6/P20/S25/L26/L30 in PsaL (Fig. 5B). Because TpLhcq10 is homologous to CgLhcr9 (30), we compared the amino acid sequence of Lhcq10 with the Lhcq and Lhcr subfamilies in T. pseudonana (Fig. S10A, B). It was shown that the amino acid sequence of L126–W155 in Lhcq10 is not conserved in the Lhcq subfamily consisting of Lhcq1, Lhcq2, Lhcq3, Lhcq4, Lhcq5, Lhcq6, Lhcq7, Lhcq8, and Lhcq9 (Fig. S10A), as well as in the Lhcr subfamily consisting of Lhcr1, Lhcr3, Lhcr4, Lhcr7, Lhcr11, Lhcr12, Lhcr14, Lhcr17, Lhcr18, Lhcr19, and Lhcr20 according to Kumazawa et al. (30) (Fig. S10B).

Structural characteristics of FCPI-3, 4, and 5.

(A) Interactions among FCPIs and between FCPIs and PsaL viewed from the stromal (left) and lumenal (right) sides. The areas encircled by black squares are enlarged in panels B–D. The PSI and FCPI subunits are colored grey and yellow, respectively. Protein-protein interactions are shown as different colors: blue, FCPI-3; magenta, FCPI-4; orange, FCPI-5; purple, PsaL. (B–D) Protein-protein interactions between FCPI-3 and PsaL (B), between FCPI-4 and FCPI-5 (C), and between FCPI-5 and PsaL (D), respectively. Interactions are indicated by dashed lines, and the numbers are distances in Å. Amino acid residues participating in the interactions are labeled; for example, L126/3 means Leu126 of FCPI-3. 4, FCPI-4; 5, FCPI-5; L, PsaL.

FCPI-4 (Lhcf10) is located near FCPI-5 through protein-protein interactions with it at distances of 2.6–3.6 Å at the lumenal side (Fig. 5A, C). The amino acid residues of Y196/P198/F199 in FCPI-4 interact with F82/F86/G87 in FCPI-5 (Fig. 5C). The amino acid sequence of Y196–F199 in Lhcf10 is not conserved in the Lhcf subfamily comprising of Lhcf1, Lhcf2, Lhcf3, Lhcf4, Lhcf5, Lhcf6, Lhcf7, Lhcf8, Lhcf9, Lhcf11, and Lhcf12 according to Kumazawa et al. (30) (Fig. S11).

FCPI-5 (Lhcq8) is located near PsaL and FCPI-4 through protein-protein interactions with them at distances of 2.6–4.1 Å at the stromal and lumenal sides (Fig. 5A, C, D). The amino acid residues of P108/Q109/A112/I115 in FCPI-5 are associated with F134/I137/S141 in PsaL at the lumenal side (Fig. 5D). The interactions of FCPI-5 with FCPI-4 are shown in Fig. 5C. The amino acid sequences of F82–G87 and P107–I115 in Lhcq8 are not conserved in the Lhcq subfamily consisting of Lhcq1, Lhcq2, Lhcq3, Lhcq4, Lhcq5, Lhcq6, Lhcq7, Lhcq9, and Lhcq10 according to Kumazawa et al.(30) (Fig. S12A, B).

Molecular insights into the assembly of FCPIs in diatom PSI-FCPI supercomplexes

To evaluate the molecular assembly of FCPI subunits in the T. pseudonana PSI-FCPI structure, we focused on both protein-protein interactions based on their close distances (Fig. 3–5) and amino acid residues of non-conserved region among 44 FCPs (Fig. S9–S12). This is because selective associations of FCPIs with PSI require specific amino acid residues of each FCPI. The protein-protein interactions among FCPI subunits and between FCPI and PSI subunits exist at the stromal and lumenal sides (Fig. 3–5) and appear to be recognized by unique amino acid residues of FCPIs, which are not conserved in each LHC subfamily (Fig. S9–S12). Thus, the binding and assembly of each FCPI subunit to PSI is likely determined based on amino acid sequences among 44 FCPs in T. pseudonana, especially in their loop regions.

The diatom C. gracilis has shown two different PSI-FCPI structures, one with 16 FCPI subunits (15) and the other with 24 FCPI subunits (16). This is because of alterations of antenna sizes of FCPIs in the C. gracilis PSI-FCPI supercomplexes in response to growth conditions, especially CO2 concentrations and temperatures (38). The C. gracilis PSI-FCPI structure has five FCPI subunits at the same binding sites of FCPI-1 to 5 in the T. pseudonana PSI-FCPI structure (Fig. 6A). A relationship of the Lhc genes encoding FCPs and RedCAP with the binding positions of FCPI-1 to 5 in T. pseudonana and C. gracilis is summarized in Fig. 6B. It should be noted that the gene names of the C. gracilis FCPIs are based on Kumazawa et al. (30) as discussed in our recent study (14).

Comparisons of structures and sequences of FCPIs in the PSI-FCPI structures between T. pseudonana and C. gracilis.

(A) Superposition of the PSI-FCPI structures between T. pseudonana and C. gracilis (PDB: 6LY5). The T. pseudonana and C. gracilis FCPI subunits are colored red and cyan, respectively. The structures are viewed from the stromal side. The FCPI-1 to 5 sites are labeled. (B) Correlation of the names of FCPIs in the structures with their genes between T. pseudonana and C. gracilis. The genes of FCPIs are derived from Kumazawa et al. (30) and Kato et al. (14) for C. gracilis. (C) Phylogenetic analysis of FCPs and RedCAPs from T. pseudonana (Tp) and C. gracilis (Cg). In addition to the RedCAP subfamily, 44 TpFCPs and 46 CgFCPs are grouped into five Lhc subfamilies and CgLhcr9 homologs. Maroon, RedCAP subfamily; magenta, Lhcq subfamily; red, Lhcz subfamily; orange, Lhcr subfamily; brown, CgLhcr9 homologs; green, Lhcf subfamily; blue, Lhcx subfamily. The FCPs and RedCAPs located at the FCPI-1 to 5 sites are labeled. The tree was inferred by IQ-TREE 2 (58) using the Q.pfam + R4 model selected with ModelFinder (59). The light purple circular symbols on the tree represent bootstrap support (%).

Phylogenetic analysis clearly showed that at the sites of FCPI-1, 2, 3, and 5 in the T. pseudonana PSI-FCPI structure, TpRedCAP, TpLhcr3, TpLhcq10, and TpLhcq8 are orthologous to CgRedCAP, CgLhcr1, CgLhcr9, and CgLhcq12, respectively (Fig. 6C). The characteristic protein loops of TpRedCAP and CpRedCAP appear to be involved in interactions with PSI at the FCPI-1 site, as described above (Fig. S8) At the FCPI-2 site, comparative analyses revealed that the amino acid residues facilitating interactions between TpLhcr3 and TpPsaB/TpPsaM closely parallel those observed in the CgLhcr1-CgPsaB and CgLhcr1-CgPsaM pairs (Fig. S13). Similarly, a high degree of similarity characterized the residues involved in the interaction pairs of TpLhcq10-TpPsaL/CgLhcr9-CgPsaL at the FCPI-3 site (Fig. S14A, B) and TpLhcq8-TpPsaL/CgLhcq12-CgPsaL at the FCPI-5 site (Fig. S14C, D). However, TpLhcf10 was not homologous to CgLhcf3 (Fig. 6C), both of which are located at the FCPI-4 site in each PSI-FCPI structure (Fig. 6A). Thus, the two diatoms appear to possess both a conserved mechanism of protein-protein interactions across characteristic protein motifs between FCPI and PSI subunits and a different interaction mechanism among FCPIs.

It is of note that the PSI-FCPI structure of C. gracilis binds much more FCPI subunits than that of T. pseudonana, namely, 16 or 24 subunits in C. gracilis found in the previous studies (15, 16) vs. 5 subunits in T. pseudonana found in the present study. We do not know the reason for this difference at present. It is possible that some of the FCPI subunits are released during detergent solubilization in T. pseudonana, but they are reserved in C. gracilis. Alternatively, it is also possible that the number of FCPI subunits is inherently less than that of C. gracilis probably due to the differences in their living environments. Further studies will clarify this question.

Extension to molecular assembly of PSI-LHCI supercomplexes

The mechanisms of protein-protein interactions for diatom PSI-FCPI supercomplexes are likely developed by specific bindings of FCPs selected from 44 TpFCPs and 46 CgFCPs in addition to RedCAPs. Like a lock-and-key mechanism, one FCP cannot be substituted by the other FCPs to form the PSI-FCPI supercomplexes in the two diatoms, e.g., TpLhcq10 cannot bind at the FCPI-2 site. The selective binding mechanism of FCPIs may dictate the fate of the molecular assembly of PSI-FCPI. It is important to note that the selective bindings of FCPIs are found for the first time by comparing structures of PSI-FCPI supercomplexes and amino acid sequences of FCPIs in the two diatom species. This approach can be extended to the LHC protein superfamily in the green and red lineages, so that comparing protein structures and sequences of PSI-LHCI supercomplexes among closely related species lays the foundation for elucidating the general mechanism of molecular assembly of PSI-LHCI supercomplexes. Thus, this study will shed light on answering the evolutionary question as to how LHCIs recognize their binding sites to PSI in photosynthetic organisms.

Methods

Cell growth and preparation of thylakoid membranes

The marine centric diatom, T. pseudonana CCMP1335, was grown in artificial seawater containing sodium metasilicate and KW21 (39) at 20 °C at a photosynthetic photon flux density of 30 μmol photons m−2 s−1 provided by white LED with bubbling of air containing 3% (v/v) CO2. The cells were harvested by centrifugation, and then disrupted by agitation with glass beads (40), followed by centrifugation to pellet the thylakoid membranes. The thylakoid membranes thus prepared were suspended with a 50 mM Mes-NaOH (pH 6.5) buffer containing 1 M betaine and 1 mM EDTA.

Purification of the PSI-FCPI supercomplex

Thylakoid membranes were solubilized with 1% (w/v) n-dodecyl-β-D-maltoside (β-DDM) at a Chl concentration of 0.5 mg mL−1 for 20 min on ice in the dark with gentle stirring. After centrifugation at 162,000 × g for 20 min at 4°C, the resultant supernatant was loaded onto a Q-Sepharose anion-exchange column (1.6 cm of inner diameter and 25 cm of length) equilibrated with a 20 mM MES-NaOH (pH 6.5) buffer containing 0.2 M trehalose, 5 mM CaCl2, and 10 mM MgCl2, and 0.03% β-DDM (buffer A). The column was washed with buffer A until the eluate became colorless. Elution was performed at a flow rate of 1.0 mL min−1 using a linear gradient of buffer A and buffer B (buffer A plus 500 mM NaCl): 0–600 min, 0–60% buffer B; 600–800 min, 60–100% buffer B; 800–900 min, 100% buffer B. The fraction enriched in PSI-FCPI was eluted at 194–247 mM NaCl, which was collected and loaded onto a linear gradient containing 10–40% (w/v) trehalose in a medium of 20 mM MES-NaOH (pH 6.5), 5 mM CaCl2, 10 mM MgCl2, 100 mM NaCl, and 0.03% β-DDM. After centrifugation at 154,000 × g for 18 h at 4°C (P40ST rotor; Hitachi), a green fraction (Fig. S1A) was collected and concentrated using a 150 kDa cut-off filter (Apollo; Orbital Biosciences) at 4,000 × g. The concentrated samples were stored in liquid nitrogen until use.

Biochemical and spectroscopic analyses of the PSI-FCPI supercomplex

The polypeptide bands of PSI-FCPI were analyzed by SDS-polyacrylamide gel electrophoresis (PAGE) containing 16% (w/v) acrylamide and 7.5 M urea according to Ikeuchi and Inoue (41) (Fig. S1B). The PSI-FCPI supercomplexes (4 µg of Chl) were solubilized by 3% lithium lauryl sulfate and 75 mM dithiothreitol for 10 min at 60°C, and loaded onto the gel. A standard molecular weight marker (SP-0110; APRO Science) was used. The absorption spectrum of PSI-FCPI was measured at room temperature using a UV-Vis spectrophotometer (UV-2450; Shimadzu) (Fig. S1C), and the fluorescence-emission spectrum of PSI-FCPI was measured at 77 K upon excitation at 430 nm using a spectrofluorometer (RF-5300PC; Shimadzu) (Fig. S1D). The pigment composition of PSI-FCPI was analyzed by HPLC according to Nagao et al. (42), and the elution profile was monitored at 440 nm (Fig. S1E).

Cryo-EM data collection

A 3-μL aliquot of the T. pseudonana PSI-FCPI supercomplex (3.0 mg of Chl mL−1) in a 20 mM MES-NaOH (pH 6.5) buffer containing 0.5 M betaine, 5 mM CaCl2, 10 mM MgCl2, and 0.03% β-DDM was applied to Quantifoil R1.2/1.3 Cu 300 mesh grids in the chamber of FEI Vitrobot Mark IV (Thermo Fisher Scientific). Then, the grid was blotted with a filter paper for 4 sec at 4°C under 100% humidity and plunged into liquid ethane cooled by liquid nitrogen. The frozen grid was transferred into a CRYO ARM 300 electron microscope (JEOL) equipped with a cold-field emission gun operated at 300 kV. All image stacks were collected from 5 × 5 holes per stage adjustment to the central hole and image shifts were applied to the surrounding holes while maintaining an axial coma-free condition. The images were recorded with an in-column energy filter with a slit width of 20 eV and at a nominal magnification of × 60,000 on a direct electron detector (Gatan K3, AMETEK). The nominal defocus range was −1.8 to −1.2 μm. Physical pixel size corresponded to 0.752 Å. Each image stack was exposed at a dose rate of 21.46 e−Å−2sec−1 for 2.33 sec in CDS mode with dose-fractionated 50 movie frames. In total 8,950 image stacks were collected.

Cryo-EM image processing

The resultant movie frames were aligned and summed using MotionCor2 (43) to yield dose-weighted images. Estimation of the contrast transfer function (CTF) was performed using CTFFIND4 (44). All of the following processes were performed using RELION-4.0 (45). In total 2,733,572 particles were automatically picked up and used for reference-free 2D classification. Then, 1,132,721 particles were selected from good 2D classes and subsequently subjected to 3D classification without any symmetry. An initial model for the first 3D classification was generated de novo from 2D classification. As shown in Fig. S2C, the final PSI-FCPI structure was reconstructed from 75,667 particles. The overall resolution of the cryo-EM map was estimated to be 2.30 Å by the gold-standard FSC curve with a cut-off value of 0.143 (Fig. S3A) (46). Local resolutions were calculated using RELION (Fig. S3C).

Model building and refinement

Two types of the cryo-EM maps were used for the model building of the PSI-FCPI supercomplex: one was a postprocessed map, and the other was a denoised map using Topaz version 0.2.4 (47). The postprocessed map was denoised using the trained model in 100 epochs with two half-maps. Initial models of each subunit in the PSI-FCPI supercomplex were built by ModelAngelo (37), and then their structures were inspected and manually adjusted against the maps with Coot (48). Each model was built based on interpretable features from the density maps at a contour level of 2.5 σ in the denoised and postprocessed maps. For the assignment of Chls, Chls a and c were distinguished by inspection of the density map corresponding to the phytol chain with the above thresholds, which was found to be the least level not to link the map of Chls with that of noise. All Chls c were assigned as Chl c1, because of difficulty in distinction between Chl c1 and Chl c2 at the present resolution. For the assignment of Cars, Fx and Ddx were distinguished based on the density covering the head group of carotenoids with the above threshold. The PSI-FCPI structure was refined with phenix.real_space_refine (49) and Servalcat (50) with geometric restraints for the protein-cofactor coordination. The final model was validated with MolProbity (51), EMRinger (52), and Q-score (53). The statistics for all data collection and structure refinement are summarized in Table S1, S2. All structural figures were made by PyMOL (54), UCSF Chimera (55), and UCSF ChimeraX (56).

Since the numbering of Chls, Cars, and other cofactors in this paper were different from those of the PDB data, we listed the relationship of their numbering in this paper with those in the PDB data in Table S6–S8.

Phylogenetic analysis

Amino acid sequences were aligned using MAFFT L-INS-i v7.490 or MAFFT E-INS-i v7.520 (57). The alignment was trimmed using ClipKit v1.4.1 with smart-gap mode. The phylogenetic tree was inferred using IQ-TREE 2 (58) with the model selected by ModelFinder (59). The tree was visualized by iTOL v6 (60). Ultrafast bootstrap approximation was performed with 1000 replicates (61).

Data availability

Atomic coordinate and cryo-EM maps for the reported structure have been deposited in the Protein Data Bank under an accession code 8XLS [https://www.rcsb.org/structure/8XLS] and in the Electron Microscopy Data Bank under an accession code EMD-38457 [https://www.ebi.ac.uk/emdb/EMD-38457].

eLife Assessment

This important study reports a high-resolution cryoEM structure of the supercomplex between photosystem I (PSI) and fucoxanthin chlorophyll a/c-binding proteins (FCPs) from the model diatom Thalassiosira pseudonana CCMP1335, with subunits, protein:protein interactions and pigments not previously seen in other diatoms or red/green photosynthetic lineages. Combining structural, sequence, and phylogenetic analyses, the authors provide convincing evidence of conserved motifs crucial for the binding of FCPs, leading to interesting speculation about the mechanisms governing the assembly of PSI-FCP supercomplexes in diatoms and their implications for related PSI-LHC supercomplexes in plants. The findings set the basis for functional experiments that will further advance the fields of photosynthesis, bioenergy, ocean biogeochemistry, and understanding of evolutionary relationships between photosynthetic organisms.

Reviewer #1 (Public Review):

The authors present the cryo-EM structure of of PSI-fucoxanthin chlorophyll a/c-binding proteins (FCPs) supercomplex from the diatom Thalassiosira pseudonana CCMP1335 at a global resolution of 2.3 Å. This exceptional resolution allows the authors to construct a near-atomic model of the entire supercomplex and elucidate the molecular details of FCPs arrangement. The high-resolution structure reveals subunits not previously identified in earlier reconstructions and models, as well as sequence analysis of PSI-FCPIs from other diatoms and red algae. Additionally, the authors use their model in conjunction with a phylogenetic analysis to compare and contrast the structural features of the T. pseudonana supercomplex with those of Chaetoceros gracilis, uncovering key structural features that contribute to the efficiency of light energy conversion in diatoms.

The study employs the advanced technique of single particle cryo-electron microscopy to visualize the complex architecture of the PSI supercomplex at near-atomic resolution and analyze the specific roles of FCPs in enhancing photosynthetic performance in diatoms.

Overall, the approach and data are both compelling and of high quality. The paper is well written and will be of wide interest for comprehending the molecular mechanisms of photosynthesis in diatoms. This work provides valuable insights for applications in bioenergy, environmental conservation, plant physiology, and membrane protein structural biology.

Reviewer #2 (Public Review):

Summary:

This manuscript elucidated the cryo-electron microscopic structure of a PSI supercomplex incorporating fucoxanthin chlorophyll a/c-binding proteins (FCPs), designated as PSI-FCPI, isolated from the diatom Thalassiosira pseudonana CCMP1335. Combining structural, sequence, and phylogenetic analyses, the authors provided solid evidence to reveal the evolutionary conservation of protein motifs crucial for the selective binding of individual FCPI subunits and provided valuable information about the molecular mechanisms governing the assembly and selective binding of FCPIs in diatoms.

Strengths:

The manuscript is well-written and presented clearly as well as consistently. The supplemental figures are also of high quality.

Weaknesses:

Only minor comments (provided in recommendations for authors) to help improve the manuscript.

Reviewer #3 (Public Review):

Summary:

Understanding the structure and function of the photosynthetic machinery is crucial for grasping its mode of action. Photosystem I (PSI) plays a vital role in light-driven electron transfer, which is essential for generating cellular reducing power. A primary strategy to mitigate light and environmental stresses involves incorporating peripheral light-harvesting proteins. Among various lineages, the number of LHCIs and their protein and pigment compositions differ significantly in PSI-LHCI structures. However, it is still unclear how LHCIs recognize their specific binding sites in the PSI core. This study aims to address this question by obtaining a high-resolution structure of the PSI supercomplex, including fucoxanthin chlorophyll a/c-binding proteins (FCPs), referred to as PSI-FCPI, isolated from the diatom Thalassiosira pseudonana. Through structural and sequence analyses, distinct protein-protein interactions are identified at the interfaces between FCPI and PSI subunits, as well as among FCPI subunits themselves.

Strengths:

The primary strength of this work lies in its superb isolation and structural determination, followed by clear discussion and conclusions. However, the interactions among the protein complexes and their relevance in formulating general rules are not definitively established. While efficiency is a crucial aspect, preventing damage is equally important, and currently, we cannot infer this from the provided structures.

Weaknesses:

The interactions among the protein complexes and their relevance in formulating general rules are not definitively established. While efficiency is a crucial aspect, preventing damage is equally important, and currently, we cannot infer this from the provided structures.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

The authors present the cryo-EM structure of of PSI-fucoxanthin chlorophyll a/c-binding proteins (FCPs) supercomplex from the diatom Thalassiosira pseudonana CCMP1335 at a global resolution of 2.3 Å. This exceptional resolution allows the authors to construct a near-atomic model of the entire supercomplex and elucidate the molecular details of FCPs arrangement. The high-resolution structure reveals subunits not previously identified in earlier reconstructions and models, as well as sequence analysis of PSI-FCPIs from other diatoms and red algae. Additionally, the authors use their model in conjunction with a phylogenetic analysis to compare and contrast the structural features of the T. pseudonana supercomplex with those of Chaetoceros gracilis, uncovering key structural features that contribute to the efficiency of light energy conversion in diatoms.

The study employs the advanced technique of single particle cryo-electron microscopy to visualize the complex architecture of the PSI supercomplex at near-atomic resolution and analyze the specific roles of FCPs in enhancing photosynthetic performance in diatoms.

Overall, the approach and data are both compelling and of high quality. The paper is well written and will be of wide interest for comprehending the molecular mechanisms of photosynthesis in diatoms. This work provides valuable insights for applications in bioenergy, environmental conservation, plant physiology, and membrane protein structural biology.

We thank you very much for your highly positive evaluation and comments on our manuscript.

Reviewer #2 (Public Review):

Summary:

This manuscript elucidated the cryo-electron microscopic structure of a PSI supercomplex incorporating fucoxanthin chlorophyll a/c-binding proteins (FCPs), designated as PSI-FCPI, isolated from the diatom Thalassiosira pseudonana CCMP1335. Combining structural, sequence, and phylogenetic analyses, the authors provided solid evidence to reveal the evolutionary conservation of protein motifs crucial for the selective binding of individual FCPI subunits and provided valuable information about the molecular mechanisms governing the assembly and selective binding of FCPIs in diatoms.

Strengths:

The manuscript is well-written and presented clearly as well as consistently. The supplemental figures are also of high quality.

Weaknesses:

Only minor comments (provided in recommendations for authors) to help improve the manuscript.

We thank you very much for your highly positive evaluation and comments on our manuscript.

Reviewer #3 (Public Review):

Summary:

Understanding the structure and function of the photosynthetic machinery is crucial for grasping its mode of action. Photosystem I (PSI) plays a vital role in light-driven electron transfer, which is essential for generating cellular reducing power. A primary strategy to mitigate light and environmental stresses involves incorporating peripheral light-harvesting proteins. Among various lineages, the number of LHCIs and their protein and pigment compositions differ significantly in PSI-LHCI structures. However, it is still unclear how LHCIs recognize their specific binding sites in the PSI core. This study aims to address this question by obtaining a high-resolution structure of the PSI supercomplex, including fucoxanthin chlorophyll a/c-binding proteins (FCPs), referred to as PSI-FCPI, isolated from the diatom Thalassiosira pseudonana. Through structural and sequence analyses, distinct protein-protein interactions are identified at the interfaces between FCPI and PSI subunits, as well as among FCPI subunits themselves.

Strengths:

The primary strength of this work lies in its superb isolation and structural determination, followed by clear discussion and conclusions. However, the interactions among the protein complexes and their relevance in formulating general rules are not definitively established. While efficiency is a crucial aspect, preventing damage is equally important, and currently, we cannot infer this from the provided structures.

Weaknesses:

The interactions among the protein complexes and their relevance in formulating general rules are not definitively established. While efficiency is a crucial aspect, preventing damage is equally important, and currently, we cannot infer this from the provided structures.

We thank you very much for your highly positive evaluation and comments on our manuscript. This study is aimed to decipher the interactions among different protein subunits within the PSI-FCPI supercomplex, from which we wish to draw their relevance in formulating general rules. While we agree that damage is equally important, it is unclear to us what kind of damage you are mentioning, and we consider that this may need to be treated in another publication, as we cannot elucidate everything in one paper.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

(1) Line 69: "Diatoms are one of the most important phytoplankton in aquatic environments and contribute to the primary production in the ocean remarkably." Check the sentence, something is missing.

We modified the sentence as follow:

"Diatoms are among the most essential phytoplankton in aquatic environments, playing a crucial role in the global carbon cycle, supporting marine food webs, and contributing significantly to nutrient cycling, thus ensuring the health and sustainability of marine ecosystems"

(2) Supplementary Figure 1B: The SDS-PAGE gel shows multiple bands. Do the authors know the identity of these proteins, or have they considered analyzing the bands using mass spectrometry? The band at ~17 kDa is particularly intense. Could you comment on this? Have you tried running a Native-PAGE gel?

We did not identify protein bands by MS analysis. The protein bands in the PSI-FCPI supercomplex of this diatom have been identified by Ikeda et al. 2013. The protein bands of our sample were similar to those of Ikeda et al. 2013. To explain this, we modified the sentences and cited Ikeda et al. 2013 in the revised manuscript (lines 89-91).

"The PSI-FCPI supercomplexes were purified from the diatom T. pseudonana CCMP1335 and analyzed by biochemical and spectroscopic techniques (Fig. S1). Notably, the protein bands of PSI-FCPI closely resembled those reported in a previous study (31)."

The ~17 kDa protein band appears to be FCPIs, which was identified in Ikeda et al. 2013. We did not perform BN-PAGE of this sample; however, we performed trehalose density gradient centrifugation (Fig. S1A).

(3) Can the authors comment on the position of the FCPI subunits in the PSI supercomplex in diatoms compared to the arrangement of LHCIs in complex with PSI in cyanobacteria, green algae, and angiosperms? This information would be useful to incorporate into the text.

We previously compared the PSI-FCPI structures of the diatom C. gracilis to the PSI-LHCI structures of land plant, green alga, and red alga (Nagao et al., 2020). Also, Xu et al. 2020 compared the C. gracilis PSI-FCPI structure to the PSI-LHCI structures of land plant, green alga, and red alga. The binding sites between FCPIs and LHCIs are conserved to some extent. However, our recent study revealed that no orthologous relationship exists among LHCs bound to PSI between primitive red algae and diatoms (Kato et al., 2024). Consequently, we found that the information obtained from structural comparisons alone is extremely limited. To avoid misinterpretation, this study focused on comparing the structures and amino acid sequences of FCPIs between T. pseudonana and C. gracilis.

(4) Line 104: Despite achieving high resolution, the authors modeled only six lipid densities (the PDB model contains actually 9 lipids, you should correct it in the text). Do you believe this is due to the detergent used for purification? Can you comment on the position, identity, and potential role of the lipids within your model?

There are 6 lipids associated with the PSI core and 3 with FCP, giving rise to a total of 9 lipids. We have described it in our original text (lines 102-104 in the modified manuscript). Additionally, our structure reveals unidentified densities which likely represent lipids; they are modeled as 88 unknown lipids (UNLs). Thus, there are more lipids in the supercomplex. However, we also observed 4 β-DDM molecules (LMT) in the structure, which are used as detergents. Thus, it is possible that some lipids have dissociated and replaced by detergents. Many of the observed lipids are located between subunits, likely contributing to the stabilization of the complex.

(5) Line 111: The global resolution is very high. Why does the unknown protein have such low resolution that it was impossible to model it properly and perform de novo identification from the density map? Is it due to a lower abundance of particles with this subunit bound? Have you tried improving this with 3D classification/ focus refinement /density modification?

The Unknown subunit (UNK) is located peripherally, and its density is significantly lower compared to the neighboring subunits, which may suggest a low abundance. We applied density modification using Topaz for 3D map denoising, but the effect was minimal. As the low abundance of UNK may be the cause, 3D classification and focus refinement also had limited impact.

(6) Figure 2A: It would be useful to show the density map for the subunit together with the model, especially to demonstrate visualization of the long loop.

We added the model and map of Psa29 to Figure S4C in the revised manuscript.

(7) Given the proximity of Psa29 to PsaC, is the protein involved in electron shuttling? If so, could you comment on this? In line 131, you state that Psa29 was not found in other organisms. Can the authors speculate on the potential role of this protein in diatoms?

We have no idea about the function of Psa29 at present. However, Psa29 does not contain any cofactors, indicating no contribution of it to electron transfer reactions. To understand the function of Psa29, a deletion mutant of this gene is required for examining its functional and physiological roles in diatom photosynthesis. To explain this, we added the following sentences to the revised manuscript (lines 129-133):

"However, the functional and physiological roles of Psa29 remain unclear at present. It is evident that Psa29 does not have any pigments, quinones, or metal complexes, suggesting no contribution of Psa29 to electron transfer reactions within PSI. Further mutagenesis studies will be necessary to investigate the role of Psa29 in diatom photosynthesis."

(8) Line 163: "Among the FCPI subunits, only FCPI-1 has BCRs in addition to Fxs and Ddxs (Figure S6A). FCPI-1 is a RedCAP, which belongs to the LHC protein superfamily but is distinct from the LHC protein family (6, 7)." It would be useful if the authors could add the carotenoid model embedded in the cryoEM density map to the figure to show the features that led to modeling BCR instead of other carotenoids. Additionally, it would be helpful to include in the text why RedCAPs differ from LHCIs and their proposed role.

We added the model and map of two BCRs in FCPI-1 (RedCAP) to Figure S4F in the revised manuscript.

Phylogenetic analysis showed that RedCAPs are distinct from the LHC protein family. This has been explained in lines 163-164. Also, the functional and physiological roles of RedCAP remain unclear. To explain this, we added the sentence "; however, the functional and physiological roles of RedCAP remain unclear" to the revised manuscript (lines 164-165).

(9) Line 185: "However, it is unknown (i) whether CgRedCAP is indeed bound to the C. gracilis PSI-FCPI supercomplex and (ii) if a loop structure corresponding to the Q96-T116 loop of TpRedCAP exists in CgRedCAP." Have the authors attempted to model the protein using AlphaFold? If so, are there significant differences? Could you speculate on the absence of RedCAP in C. gracilis? Do you believe it is due to using a different detergent or related to environmental factors?

We did not model CgRedCAP using AlphaFold. Our recent study “Kato et al. 2024” proposed that CgRedCAP binds to the LHCI-1 site in the PSI-FCPI structure based on sequence comparison. There are two types of PSI-FCPI supercomplexes, one having 16 FCPIs and the other having 24 FCPs, from C. gracilis. The different antenna sizes may depend on the growth conditions of C. gracilis (Nagao et al. 2020). These explanations were already described in the manuscript (lines 243-246).

(10) Line 193: Figure 8 is mentioned before Figures 4-7.

We are sorry for the mistake of Figure number. Figure 8 is Supplementary Figure 8, so that we modified Fig. S8B in the revised manuscript.

(11) Line 223: FCPI-4 interacts only with FCPI-5, primarily through the interaction of Y196/4 with the FCPI-5 backbone. Is this interaction facilitated by other factors such as lipids, carotenoids, or other ligands? Also, FCPI-4 occupies a peculiar position compared to other LHCIs proteins (it is peripheral to FCPI-4 and FCPI-5). Do you believe this could be due to a transient interaction with the complex? Could the presence of this protein be related to the growth conditions experienced by the plant? Are there any literature reports on environmental conditions influencing FCPI arrangements? Including this information in the text would be interesting.

Y196/4 interacts with only backbones by hydrogen-bond interactions; therefore, other cofactors do not contribute to the interactions.

We do not believe that the interaction of FCPI-4 is transient; rather, this binding appears to be stable within the complex. Given that the PSI-FCPI supercomplexes were isolated by anion exchange chromatography, FCPI-4 and FCPI-5 are tightly associated within this complex. However, it is important to note that the expression of diatom FCPI proteins can indeed vary depending on growth conditions, as highlighted in our previous study (Nagao et al., 2020). While the peculiar position of FCPI-4 may not be directly related to transient interactions, environmental conditions could still influence the overall arrangement and expression levels of FCPIs. This information has already been described in the manuscript (lines 243-246).

(12) Given the high resolution of your map, the overall model quality does not seem to match the map quality. Specifically, the clash score (10) and sidechain outliers (3%) are elevated. Could you comment on this? Do you believe it is related to the high number of ligands?

Our structure contains a total of 295 ligands, including cofactors, detergents, and unknown lipids. We believe the high clash score and number of sidechain outliers are due to the large number of ligands present.

(13) Supplementary Figure 2: You should show the 3D classes that were discarded.

According to your comment, we added the 3D classes that were discarded and the sentence "Red boxes highlight selected particles from each 3D classification." to Figure S2 and its legend in the revised manuscript.

(14) Which masks were used for refinement? How were they generated, and which parameters were chosen? This information should be added to the Materials and Methods section. You should show the masks used during classification, for example.

We used a 240 Å spherical mask for refinement and classification, without applying any reference mask as input. To explain this, we added the corresponding sentence to Methods in the revised manuscript (lines 347-348) as follow:

"A 240-Å spherical mask was used during the 3D classification and refinement processes."

(15) Were any extra proteins detected in the early stages of the cryoEM analysis (i.e., 2D classification) that were discarded? Could you visualize the superior oligomeric states of the supercomplex?

In the single-particle analysis, no larger particles than the analyzed complex were detected. The results of 2D classification using a sufficiently large spherical mask with a diameter of 320 Å are shown below.

(16) Have you tried using cryoSPARC for data analysis? If so, could you comment on that?

We did not use cryoSPARC for data analysis.

Reviewer #2 (Recommendations For The Authors):

I have some minor comments below to help improve the manuscript. The line numbers below refer to those in the Word version of the manuscript.

(1) Figure 1 legend, line 559, "membrane normal"? Panel A and B, structures with the same colors, do they refer to the closely related or interacted parts? For example, the red color for FCP1-1 in A and PsaA in B. If not, the authors may want to clarify it.

The term 'membrane normal' refers to the direction perpendicular to the surface of a membrane. It is a concept frequently used in physics and biology to describe the orientation relative to the membrane's plane.

We do not refer to either the closely related or interacted parts used in Figure 1. According to your comments, the colors of subunits were revised in the revised manuscript.

(2) Line 109-117. "Psa28 is a novel subunit found in the C. gracilis PSI-FCPI structure, and its name follows the nomenclature as suggested previously (31).... After psaZ, the newly identified genes should be named psa27, psa28, etc., and the corresponding proteins are called Psa27, Psa28, etc... Psa28 was also named PsaR in the PSI-FCPI structure of C. gracilis (16)". It is confusing. Was Psa28 named twice, PsaR and Psa28? It would be helpful to add a simple explanation here.

According to your comment, we modified the sentence as follow (lines 117-118):

" However, Xu et al. named the subunit as PsaR in the PSI-FCPI structure of C. gracilis "

(3) Line 134, "One of the Car molecules in PsaJ was identified as ZXT103 in the T. pseudonana PSI-FCPI structure but it is BCR112 in the C. gracilis PSI-FCPI structure (15)". Figure S4D mentioned BCR863 but did not mention BCR112. Figure S4C, D, it may need better explanations of the colors and labels, and indicate which parts are from T. pseudonana or C. gracilis.

BCR112 was misnumbered; the correct number is BCR103. In response to your comments, we revised Figure S4C and D by labeling the characteristic pigments in the revised manuscript.

(4) Figure S7, although mentioned in the legend, it would be helpful to label interaction pairs on the figure directly with corresponding colours.

According to your comments, we modified the Figure and legends in the revised manuscript.

(5) Figure 3E, it is better to avoid red/green colours in one figure as some readers may be colour-blind. It would also be helpful to label each FCPI with the same colour as its structure on the figure directly.

According to your comments, we modified Figure 3E in the revised manuscript.

(6) Line 185, "structures similar to the Q96-T116 loop in TpRedCAP found in the present study (Figure 8B).". The authors refer to Figure S8B? I have the same comment for line 186, Figure 8C.

We are sorry for the mistake of Figure number. Figure 8 is Supplementary Figure 8, so we modified it as Fig. S8B in the revised manuscript.

(7) Line 270, "TpLhcq10 cannot bind at the FCPI-2 site". Why not use FCPI-3 for TpLhcq10?

This means that the gene product of TpLhcq10 binds at the FCPI-3 site but not at the other sites such as FCPI-2. To avoid misreading, we modified the sentence as follows:

"TpLhcq10 binds specifically at the FCPI-3 site but not at the other sites such as FCPI-2" (lines 278-279)

Reviewer #3 (Recommendations For The Authors):

I have no technical or conceptual suggestions at the current stage.

Thank you.

Version 2

Abstract

Abstract

Photosynthetic organisms exhibit remarkable diversity in their light-harvesting complexes (LHCs). LHCs are associated with photosystem I (PSI), forming a PSI-LHCI supercomplex. The number of LHCI subunits, along with their protein sequences and pigment compositions, has been found to differ greatly among the PSI-LHCI structures. However, the mechanisms by which LHCIs recognize their specific binding sites within the PSI core remain unclear. In this study, we determined the cryo-electron microscopy structure of a PSI supercomplex incorporating fucoxanthin chlorophyll a/c-binding proteins (FCPs), designated as PSI-FCPI, isolated from the diatom Thalassiosira pseudonana CCMP1335. Structural analysis of PSI-FCPI revealed five FCPI subunits associated with a PSI monomer; these subunits were identified as RedCAP, Lhcr3, Lhcq10, Lhcf10, and Lhcq8. Through structural and sequence analyses, we identified specific protein-protein interactions at the interfaces between FCPI and PSI subunits, as well as among FCPI subunits themselves. Comparative structural analyses of PSI-FCPI supercomplexes, combined with phylogenetic analysis of FCPs from T. pseudonana and the diatom Chaetoceros gracilis, underscore the evolutionary conservation of protein motifs crucial for the selective binding of individual FCPI subunits. These findings provide significant insights into the molecular mechanisms underlying the assembly and selective binding of FCPIs in diatoms.

Introduction

Oxygenic photosynthesis in cyanobacteria, algae, and land plants converts solar energy into chemical energy and releases molecular oxygen into the atmosphere (1). The conversion of light energy takes place within two multi-subunit membrane protein complexes, known as photosystem I (PSI) and photosystem II (PSII), which perform light harvesting, charge separation, and electron transfer reactions (2–5). To optimize light energy capture, numerous light-harvesting antenna subunits are associated with the periphery of the PSI and PSII core complexes, transferring excitation energy to the respective photosystem cores (1). These light-harvesting antennae exhibit significant diversity among photosynthetic organisms, both in protein sequences and pigment compositions, and can be broadly categorized into two major groups: membrane proteins and water-soluble proteins (1).

The membrane protein category primarily consists of the light-harvesting complex (LHC) protein superfamily (6, 7), which absorbs light energy through chlorophylls (Chls) and carotenoids (Cars). The number and types of Chls and Cars vary significantly among LHCs, which can be grouped into green and red lineages, leading to color diversity in photosynthetic organisms (8). The green lineage includes green algae and land plants, while the red lineage encompasses red algae, diatoms, haptophytes, cryptophytes, and dinoflagellates (8). LHCs specific to PSI (LHCIs) bind to a eukaryotic PSI monomer, forming a PSI-LHCI supercomplex (9, 10), the structures of which have been revealed by cryo-electron microscopy (cryo-EM) in various eukaryotes (9, 10). In the red lineage, the number of LHCIs and their protein sequences and pigment compositions exhibit considerable variation among the PSI-LHCI structures of red algae (11–14), a diatom (15, 16), a cryptophyte (17), and dinoflagellates (18, 19).

Recently, we demonstrated the conservation and diversity of LHCIs among red-lineage algae through structural and phylogenetic analyses of PSI-LHCI supercomplexes (14). This study revealed that while the binding sites of LHCIs to PSI were conserved to some extent among red-lineage algae, their evolutionary relationships were weak. It is known that LHCIs have similar overall protein structures across photosynthetic organisms, with particular similarity in their three-transmembrane helices, regardless of whether they belong to the green or red lineages (9, 10). However, individual LHCIs have altered their sequences and structures to adapt their respective binding sites to the PSI cores during the assembly of PSI-LHCI supercomplexes. These observations raise a critical question: how do LHCIs recognize their binding sites in the PSI core?

Diatoms are among the most essential phytoplankton in aquatic environments, playing a crucial role in the global carbon cycle, supporting marine food webs, and contributing significantly to nutrient cycling, thus ensuring the health and sustainability of marine ecosystems (20). Diatoms possess unique LHCs known as fucoxanthin Chl a/c-binding proteins (FCPs), which differ in pigment composition and amino acid sequences from the LHCs of land plants (21–23). Previous studies have reported the isolation and structural characterization of PSI-FCPI supercomplexes isolated from the diatom Chaetoceros gracilis (15, 16, 24–29). Kumazawa et al. showed significant diversity in FCPs between C. gracilis and Thalassiosira pseudonana, with 46 and 44 FCPs identified, respectively (30). These FCPs are categorized into multiple, closely related subgroups (30), and their amino acid sequences are not entirely identical between the two diatoms. Consequently, comparing FCPIs, including their amino acid residues and protein structures at similar binding sites in PSI-FCPIs, may provide molecular insights into how FCPIs interact with PSI. However, an overall structure of the T. pseudonana PSI-FCPI supercomplex has yet to be solved.

In this study, we solved the structure of the PSI-FCPI supercomplex from T. pseudonana CCMP1335 at a resolution of 2.30 Å by cryo-EM single-particle analysis. The structure reveals a PSI-monomer core and five FCPI subunits. Structural and sequence comparisons highlight unique protein-protein interactions between each FCPI subunit and PSI. Based on these findings, we discuss the molecular assembly and selective binding mechanisms of FCPI subunits in diatom species.

Results and discussion

Overall structure of the T. pseudonana PSI-FCPI supercomplex

The PSI-FCPI supercomplexes were purified from the diatom T. pseudonana CCMP1335 and analyzed by biochemical and spectroscopic techniques (Fig. S1). Notably, the protein bands of PSI-FCPI closely resembled those reported in a previous study (31). Cryo-EM images of the PSI-FCPI supercomplex were obtained using a JEOL CRYO ARM 300 electron microscope operated at 300 kV. The final cryo-EM map was determined at a resolution of 2.30 Å with a C1 symmetry (Fig. S2, S3, Table S1), based on the “gold standard” Fourier shell correlation (FSC) = 0.143 criterion (Fig. S3A).

The atomic model of PSI-FCPI was built based on the cryo-EM map obtained (see Methods; Fig. S3, Table S1–S3). The structure reveals a monomeric PSI core associated with five FCPI subunits (Fig. 1A, B). The five FCPI subunits were named FCPI-1 to 5 (Fig. 1A), following the nomenclature of LHCI subunits in the PSI-LHCI structure of Cyanidium caldarium RK-1 (NIES-2137) (14). Specifically, the positions of FCPI-1 and FCPI-2 in the T. pseudonana PSI-FCPI structure (Fig. 1A) correspond to those of LHCI-1 and LHCI-2 in the C. caldarium PSI-LHCI structure. The PSI core comprises 94 Chls a, 18 β-carotenes (BCRs), 1 zeaxanthin (ZXT), 3 [4Fe-4S] clusters, 2 phylloquinones, and 6 lipid molecules, whereas the five FCPI subunits include 45 Chls a, 7 Chls c, 2 BCRs, 15 fucoxanthins (Fxs), 7 diadinoxanthins (Ddxs), and 3 lipid molecules (Table S3).

Overall structure of the PSI-FCPI supercomplex from T. pseudonana. Structures are viewed from the stromal side (left panels) and from the direction perpendicular to the membrane normal (right panels). Only protein structures are depicted, with cofactors omitted for clarity. The FCPI (A) and PSI-core (B) subunits are labeled and colored distinctly. The five FCPI subunits are labeled as FCPI-1 to 5 (red), with their corresponding gene products indicated in parentheses (black) in panel A.

Structure of the T. pseudonana PSI core

The PSI core contains 12 subunits, 11 of which are identified as PsaA, PsaB, PsaC, PsaD, PsaE, PsaF, PsaI, PsaJ, PsaL, PsaM, and Psa29 (Fig. 1B). The remaining subunit could not be assigned due to insufficient map resolution and was therefore modeled as polyalanines (Fig. S4A). This unidentified subunit, designated as Unknown, occupies the same site as Psa28 in the C. gracilis PSI-FCPI (15). The structural comparison reveals that Unknown closely resembles Psa28 in the C. gracilis PSI-FCPI (Fig. S4B). Psa28, a novel subunit identified in the C. gracilis PSI-FCPI structure (15), follows the previously established nomenclature rule (32). Historically, genes encoding PSI proteins have been designated as psaA, psaB, and so forth. PsaZ was identified in the PSI cores of Gloeobacter violaceus PCC 7421 (33, 34). Subsequent discoveries led to the designation of a new subunit as Psa27, which was identified in the PSI cores of Acaryochloris marina MBIC11017 (35–37). Consequently, we designated this novel subunit as Psa28 (15). However, Xu et al. referred to this subunit as PsaR in the PSI-FCPI structure of C. gracilis (16).

Psa29 is newly identified in the T. pseudonana PSI-FCPI structure using ModelAngelo (38) and the NCBI database (https://www.ncbi.nlm.nih.gov/) (Fig. 2). The subunit corresponding to Psa29 was also observed previously in the C. gracilis PSI-FCPI structures (15, 16), where it was modeled as polyalanines and referred to as either Unknown1 (15) or PsaS (16). Psa29 exhibits a unique structure distinct from the other PSI subunits in the T. pseudonana PSI-FCPI (Fig. 2A, Fig. S4C) and engages in multiple interactions with PsaB, PsaC, PsaD, and PsaL at distances of 2.5–3.2 Å (Fig. 2B–G). Sequence analyses suggest that Psa29 has undergone evolutionary divergence between Bacillariophyceae (diatoms) and Bolidophyceae, the latter of which is a sister group of diatoms within Stramenopiles (Fig. 2H), although this subunit has not been found in other organisms. The arrangement of PSI subunits in the T. pseudonana PSI-FCPI is virtually identical to that in the C. gracilis PSI-FCPI structures already reported (15, 16). However, the functional and physiological roles of Psa29 remain unclear at present. It is evident that Psa29 does not have any pigments, quinones, or metal complexes, suggesting no contribution of Psa29 to electron transfer reactions within PSI. Further mutagenesis studies will be necessary to investigate the role of Psa29 in diatom photosynthesis.

Structure and diversity of Psa29. (A) Structure of Psa29 depicted as cartoons. Psa29 was modeled from V47 to L178. (B) Cryo-EM map of Psa29 and its surrounding environment, viewed from the stromal side. The red-squared areas are enlarged in panels C–G. Yellow, PsaB; cyan, PsaC; blue, PsaD; magenta, PsaL; orange, Psa29. (C–G) Protein-protein interactions of Psa29 with PsaB/PsaC (C), PsaC/PsaD (D), PsaB/PsaD (E), PsaB/PsaD/PsaL (F), and PsaD (G). Interactions are indicated by dashed lines, and the numbers are distances in Å. Amino acid residues participating in the interactions are labeled; for example, A71/C indicates Ala71 of PsaC. B, PsaB; C, PsaC; D, PsaD, L, PsaL; 29, Psa29. (H) Phylogenetic analysis of Psa29 in photosynthetic organisms. A maximum-likelihood tree of Psa29 proteins was inferred using IQ-TREE v2.2.2.7 with the WAG+F+G4 model and a trimmed alignment of 22 sequences comprising 245 amino acid residues. Numbers at the nodes represent ultrafast bootstrap support (%) (1000 replicates). The tree was mid-point rooted between diatoms and Bolidophyceae Parmales. Psa29 of T. pseudonana CCMP1335 is indicated by a red underline.

The number and arrangement of Chls and Cars within the PSI core in the T. pseudonana PSI-FCPI structure (Fig. S4D, E) are largely similar to those in the C. gracilis PSI-FCPI structure (15). However, Chl a102 of PsaI was found in the T. pseudonana PSI-FCPI structure but not in the C. gracilis PSI-FCPI structure (15), whereas a844 of PsaA and BCR843 of PsaB were identified in the C. gracilis PSI-FCPI structure (15) but not in the T. pseudonana PSI-FCPI structure. One of the Car molecules in PsaJ was identified as ZXT103 in the T. pseudonana PSI-FCPI structure, while it is BCR103 in the C. gracilis PSI-FCPI structure (15).

Structure of the T. pseudonana FCPIs

Kumazawa et al. classified 44 Lhc genes in T. pseudonana, designating them as Lhcf, Lhcq, Lhcr, Lhcx, Lhcz, and CgLhcr9 homologs (30). Based on this classification, the five FCPI subunits in the PSI-FCPI structure were identified using five genes: RedCAP, Lhcr3, Lhcq10, Lhcf10, and Lhcq8, corresponding to FCPI-1 to 5, respectively (Fig. 1A). It is important to note that RedCAP is not included among the 44 Lhc genes (30) but is classified within the LHC protein superfamily (6, 7). For the assignment of each FCPI subunit, we focused on characteristic amino acid residues derived from their cryo-EM map, especially S61/V62/Q63 in FCPI-1; A70/R71/W72 in FCPI-2; Y64/R65/E66 in FCPI-3; M63/R64/Y65 in FCPI-4; and A62/R63/R64 in FCPI-5 (Fig. S5). The root mean square deviations (RMSDs) of the structures between FCPI-4 and the other four FCPIs range from 1.91 to 3.73 Å (Table S4).

Each FCPI subunit binds several Chl and Car molecules: 7 Chls a/1 Chl c/2 Fxs/3 Ddxs/2 BCRs in FCPI-1; 10 Chls a/1 Chl c/3 Fxs/1 Ddx in FCPI-2; 7 Chls a/3 Chls c/2 Fxs/2 Ddxs in FCPI-3; 11 Chls a/2 Chls c/4 Fxs in FCPI-4; and 10 Chls a/4 Fxs/1 Ddx in FCPI-5 (Fig. S6A–E, Table S3). The axial ligands of the central Mg atoms of Chls within each FCPI are primarily provided by the main and side chains of amino acid residues (Table S5). Potential excitation-energy-transfer pathways can be proposed based on the close physical interactions among Chls between FCPI-3 and PsaA, between FCPI-3 and PsaL, between FCPI-1 and PsaI, and between FCPI-2 and PsaB (Fig. S7).

Structural characteristics of RedCAP and its evolutionary implications

Among the FCPI subunits, only FCPI-1 contains two BCRs in addition to Fxs and Ddxs (Fig. S6A, F). This is the first report of BCR binding to FCPIs in diatoms. FCPI-1 is identified as RedCAP, a member of the LHC protein superfamily but distinct from the LHC protein family (6, 7); however, the functional and physiological roles of RedCAP remain unknown. FCPI-1 is positioned near PsaB, PsaI, and PsaL through protein-protein interactions with these subunits on both the stromal and lumenal sides (Fig. 3A). At the stromal side, I138 and S139 of FCPI-1 interact with K121, G122, and F125 of PsaL (Fig. 3B), whereas at the lumenal side, multiple interactions occur between I109 of FCPI-1 and F5 of PsaI, between T105/L106/T108 of FCPI-1 and W92/P94/F96 of PsaB, and between E102/W103 of FCPI-1 and S71/I73 of PsaL (Fig. 3C). The protein-protein interactions at the lumenal side (Fig. 3C) appear to be caused by a loop structure of FCPI-1 from Q96 to T116 (pink in Fig. 3D), which is unique to FCPI-1 but absent in the other four FCPI subunits (pink in Fig. 3E). This loop structure is inserted into a cavity formed by PsaB, PsaI, and PsaL (Fig. 3C, D). These findings indicate that the Q96–T116 loop of FCPI-1 specifically recognizes and binds to the cavity provided by the PSI subunits.

Structural characteristics of FCPI-1 (RedCAP). (A) Interactions of FCPI-1 with PsaB, PsaI, and PsaL viewed from the stromal (left) and lumenal (right) sides. The areas encircled by black squares are enlarged in panels B and C. Yellow, PsaB; magenta, PsaI; dark red, PsaL; red, FCPI-1. (B, C) Protein-protein interactions of FCPI-1 with PsaL (B) and with PsaB/PsaI/PsaL (C). Interactions are indicated by dashed lines, and the numbers are distances in Å. Amino acid residues involved in the interactions are labeled; for example, S139/1 indicates Ser139 of FCPI-1. B, PsaB; I, PsaI; L, PsaL; 1, FCPI-1. (D) Characteristic loop structure from Q96 to T116 in FCPI-1, viewed from the lumenal side. Q96 and T116 are labeled with sticks, and the Q96–T116 loop is colored pink. (E) Superpositions of FCPI-1 with FCPI-2, FCPI-3, FCPI-4, and FCPI-5. Only proteins are depicted. Q96 and T116 in the Q96–T116 loop of FCPI-1 are shown with sticks.

RedCAP of C. gracilis (CgRedCAP) was not identified in the C. gracilis PSI-FCPI structures (15, 16). As previously discussed (14), we proposed that CgRedCAP may bind to the C. gracilis PSI core at a site similar to LHCI-1 in the red alga C. caldarium PSI-LHCI through sequence analysis. This site corresponds to the FCPI-1 site in the PSI-FCPI of T. pseudonana in this study. A sequence alignment between RedCAP of T. pseudonana (TpRedCAP) and CgRedCAP is shown in Fig. S8A, exhibiting a 72% sequence similarity. CgRedCAP contains a protein motif, Q106–I113 (QWGTLATI), corresponding to E102–I109 (EWGTLATI) in TpRedCAP (Fig. 3C). These findings suggest the potential binding of CgRedCAP to PSI in C. gracilis at a position similar to FCPI-1 in the T. pseudonana PSI-FCPI structure. However, it remains unclear (i) whether CgRedCAP is indeed bound to the C. gracilis PSI-FCPI supercomplex and (ii) if a loop structure corresponding to the Q96–T116 loop of TpRedCAP exists in CgRedCAP. Further structural studies of the C. gracilis PSI-FCPI are required to elucidate the molecular assembly mechanism of diatom RedCAPs.

RedCAPs have been found in the PSI-LHCI structures of the red alga Porphyridium purpureum (13) and a PSI supercomplex with alloxanthin Chl a/c-binding proteins (PSI-ACPI) in the cryptophyte Chroomonas placoidea (17), as summarized in our previous study (14). Both P. purpureum RedCAP (PpRedCAP) and C. placoidea RedCAP (CpRedCAP) exhibit loop structures similar to the Q96–T116 loop in TpRedCAP observed in the present study (Fig. S8B). Multiple sequence alignments of TpRedCAP with PpRedCAP and CpRedCAP are shown in Fig. S8C, revealing sequence similarities of 39% and 60%, respectively. PpRedCAP contains a protein motif of V105–L112 (VWGPLAQL), while CpRedCAP has a protein motif of Q117–A124 (QWGPLASA). These motifs correspond to E102–I109 (EWGTLATI) in TpRedCAP; however, the sequence conservation between TpRedCAP and PpRedCAP/CpRedCAP is lower than between TpRedCAP and CgRedCAP. Among the four RedCAPs, the amino acids Trp, Gly, Leu, and Ala are conserved in the protein motifs (xWGxLAxx), implying that this conserved loop structure contributes to the binding of RedCAP to PSI across the red-lineage algae.

Protein-protein interactions of the other FCPI subunits

FCPI-2 (Lhcr3) is positioned near PsaB and PsaM, engaging in protein-protein interactions with these subunits at distances of 3.0–4.3 Å at both the stromal and lumenal sides (Fig. 4). The amino acid residues I63/T65/D66/Y69/W134/Y138/D140 of FCPI-2 are associated with W153/L154/K159/F160/W166 of PsaB at the stromal side (Fig. 4B), while F116 and F120 of FCPI-2 interact with F5/I9/M12 of PsaM at the lumenal side (Fig. 4C). The amino acid sequences corresponding to I63–Y69, F116–F120, and W134–D140 in Lhcr3 are not conserved in the Lhcr subfamily, comprising Lhcr1, Lhcr4, Lhcr7, Lhcr11, Lhcr12, Lhcr14, Lhcr17, Lhcr18, Lhcr19, and Lhcr20, as reported by Kumazawa et al. (30) (Fig. S9).

Structural characteristics of FCPI-2. (A) Interactions of FCPI-2 with PsaB and PsaM viewed from the stromal (left) and lumenal (right) sides. The areas encircled by black squares are enlarged in panels B and C. PSI subunits are colored grey, and FCPI subunits are colored yellow. Protein-protein interactions are shown in different colors: green, FCPI-2; cyan, PsaB; pink, PsaM. (B, C) Protein-protein interactions of FCPI-2 with PsaB (B) and PsaM (C). Interactions are indicated by dashed lines, and the numbers represent distances in Å. Amino acid residues involved in the interactions are labeled; for example, Y138/2 indicates Tyr138 of FCPI-2. B, PsaB; M, PsaM; 2, FCPI-2.

FCPI-3 (Lhcq10) is positioned near PsaL, with protein-protein interactions at distances of 2.3–4.2 Å at the stromal side (Fig. 5A, B). The amino acid residues L126/I130/L142/Y146/W147/V148/W155 of FCPI-3 are associated with F4/K6/P20/S25/L26/L30 of PsaL (Fig. 5B). Given the homology between TpLhcq10 and CgLhcr9 (30), we compared the amino acid sequence of Lhcq10 with the Lhcq and Lhcr subfamilies in T. pseudonana (Fig. S10A, B). The sequence L126–W155 of Lhcq10 is not conserved in the Lhcq subfamily, comprising Lhcq1, Lhcq2, Lhcq3, Lhcq4, Lhcq5, Lhcq6, Lhcq7, Lhcq8, and Lhcq9 (Fig. S10A), nor in the Lhcr subfamily, comprising Lhcr1, Lhcr3, Lhcr4, Lhcr7, Lhcr11, Lhcr12, Lhcr14, Lhcr17, Lhcr18, Lhcr19, and Lhcr20, as reported by Kumazawa et al. (30) (Fig. S10B).

Structural characteristics of FCPI-3, 4, and 5. (A) Interactions among FCPIs and between FCPIs and PsaL, viewed from the stromal (left) and lumenal (right) sides. The areas encircled by black squares are enlarged in panels B–D. PSI subunits are colored grey, and FCPI subunits are colored yellow. Protein-protein interactions are shown in different colors: blue, FCPI-3; magenta, FCPI-4; orange, FCPI-5; purple, PsaL. (B–D) Protein-protein interactions between FCPI-3 and PsaL (B), between FCPI-4 and FCPI-5 (C), and between FCPI-5 and PsaL (D). Interactions are indicated by dashed lines, and the numbers represent distances in Å. Amino acid residues involved in the interactions are labeled; for example, L126/3 indicates Leu126 of FCPI-3. L, PsaL; 3, FCPI-3; 4, FCPI-4; 5, FCPI-5.

FCPI-4 (Lhcf10) is positioned near FCPI-5 through protein-protein interactions with it at distances of 2.6–3.6 Å at the lumenal side (Fig. 5A, C). The amino acid residues Y196/P198/F199 of FCPI-4 interact with F82/F86/G87 of FCPI-5 (Fig. 5C). The amino acid sequence Y196–F199 of Lhcf10 is not conserved in the Lhcf subfamily, comprising Lhcf1, Lhcf2, Lhcf3, Lhcf4, Lhcf5, Lhcf6, Lhcf7, Lhcf8, Lhcf9, Lhcf11, and Lhcf12, as reported by Kumazawa et al. (30) (Fig. S11).

FCPI-5 (Lhcq8) is positioned near PsaL and FCPI-4 through protein-protein interactions at distances of 2.6–4.1 Å at both the stromal and lumenal sides (Fig. 5A, C, D). The amino acid residues P108/Q109/A112/I115 of FCPI-5 interact with F134/I137/S141 of PsaL at the lumenal side (Fig. 5D). The interactions between FCPI-5 and FCPI-4 are shown in Fig. 5C. The amino acid sequences F82–G87 and P107–I115 of Lhcq8 are not conserved in the Lhcq subfamily, comprising Lhcq1, Lhcq2, Lhcq3, Lhcq4, Lhcq5, Lhcq6, Lhcq7, Lhcq9, and Lhcq10, as reported by Kumazawa et al.(30) (Fig. S12A, B).

Molecular insights into the assembly of FCPIs in diatom PSI-FCPI supercomplexes

To evaluate the molecular assembly of FCPI subunits in the T. pseudonana PSI-FCPI structure, we focused on protein-protein interactions based on their close proximities (Fig. 3–5) and the amino acid residues in non-conserved regions among 44 FCPs (Fig. S9–S12). This approach is based on the premise that selective associations of FCPIs with PSI require specific amino acid residues unique to each FCPI. Protein-protein interactions among FCPI subunits, as well as between FCPI and PSI subunits, occur at both the stromal and lumenal sides (Fig. 3–5), and are likely recognized by unique amino acid residues of FCPIs that are not conserved in each LHC subfamily (Fig. S9–S12). Thus, the binding and assembly of each FCPI subunit to PSI are likely determined by the amino acid sequences within the loop regions of the 44 FCPs in T. pseudonana.

The diatom C. gracilis exhibits two distinct PSI-FCPI structures: one with 16 FCPI subunits (15) and the other with 24 FCPI subunits (16). These structural variations arise from changes in the antenna sizes of FCPIs within the C. gracilis PSI-FCPI supercomplexes, in response to varying growth conditions, especially CO2 concentrations and temperatures (39). Notably, the C.

gracilis PSI-FCPI structure contains five FCPI subunits located at the same binding sites as FCPI-1 to 5 in the T. pseudonana PSI-FCPI structure (Fig. 6A). A summary of the relationship between the Lhc genes encoding FCPs, the distinct gene RedCAP, and the binding positions of FCPI-1 to 5 in T. pseudonana and C. gracilis is shown in Fig. 6B. The gene nomenclature for the C. gracilis FCPIs follows the conventions established by Kumazawa et al. (30), as discussed in our recent study (14).

Comparisons of structures and sequences of FCPIs in the PSI-FCPI structures between T. pseudonana and C. gracilis. (A) Superposition of the PSI-FCPI structures between T. pseudonana and C. gracilis (PDB: 6LY5). FCPI subunits from T. pseudonana and C. gracilis are colored red and cyan, respectively. The structures are viewed from the stromal side. The FCPI-1 to 5 sites are labeled. (B) Correlation of the names of FCPIs in the structures with their corresponding genes between T. pseudonana and C. gracilis. The FCPI genes are derived from Kumazawa et al. (30) and Kato et al. (14) for C. gracilis. (C) Phylogenetic analysis of FCPs and RedCAPs from T. pseudonana (Tp) and C. gracilis (Cg). In addition to the RedCAP family, 44 TpFCPs and 46 CgFCPs are grouped into five Lhc subfamilies and CgLhcr9 homologs. Maroon, RedCAP family; magenta, Lhcq subfamily; red, Lhcz subfamily; orange, Lhcr subfamily; brown, CgLhcr9 homologs; green, Lhcf subfamily; blue, Lhcx subfamily. The FCPs and RedCAPs located at the FCPI-1 to 5 sites are labeled. The tree was inferred using IQ-TREE 2 (59) with the Q.pfam + R4 model selected by ModelFinder (60). The light purple circular symbols on the tree represent bootstrap support (%).

Phylogenetic analysis clearly showed that at the FCPI-1, 2, 3, and 5 sites in the T. pseudonana PSI-FCPI structure, TpRedCAP, TpLhcr3, TpLhcq10, and TpLhcq8 are orthologous to CgRedCAP, CgLhcr1, CgLhcr9, and CgLhcq12, respectively (Fig. 6C). The characteristic protein loops of TpRedCAP and CpRedCAP likely participate in interactions with PSI at the FCPI-1 site, as noted above (Fig. S8). At the FCPI-2 site, comparative analyses revealed that the amino acid residues facilitating interactions between TpLhcr3 and TpPsaB/TpPsaM closely parallel those observed in the CgLhcr1-CgPsaB and CgLhcr1-CgPsaM pairs (Fig. S13). Similarly, a high degree of similarity characterized the residues involved in the interaction pairs of TpLhcq10-TpPsaL/CgLhcr9-CgPsaL at the FCPI-3 site (Fig. S14A, B), as well as TpLhcq8-TpPsaL/CgLhcq12-CgPsaL at the FCPI-5 site (Fig. S14C, D), underscoring the conserved nature of these interactions. However, TpLhcf10 was not homologous to CgLhcf3 (Fig. 6C), despite both being located at the FCPI-4 site in their respective PSI-FCPI structures (Fig. 6A). These findings suggest that the two diatoms possess both a conserved mechanism of protein-protein interactions across characteristic protein motifs between FCPI and PSI subunits, and a different mechanism of interactions among FCPIs.

It is notable that the C. gracilis PSI-FCPI structure binds remarkably more FCPI subunits than that of T. pseudonana, e.g., 16 or 24 subunits in C. gracilis as reported in the previous studies (15, 16), versus 5 subunits in T. pseudonana in the present study. The reason for this difference remains unclear. One possibility is that some FCPI subunits are released during detergent solubilization in T. pseudonana, while they are retained in C. gracilis. Alternatively, the number of FCPI subunits may be inherently lower in T. pseudonana, which may reflect adaptations to different living environments. Further studies are needed to resolve this question.

Extension to molecular assembly of PSI-LHCI supercomplexes

The mechanisms of protein-protein interactions in diatom PSI-FCPI supercomplexes are likely developed by the specific binding of FCPs selected from 44 TpFCPs and 46 CgFCPs in addition to RedCAPs. Like a lock-and-key mechanism, one FCP cannot be substituted by another in forming the PSI-FCPI supercomplexes in the two diatoms; for example, TpLhcq10 binds specifically at the FCPI-3 site but not at the other sites such as FCPI-2. This selective binding mechanism of FCPIs may dictate the molecular assembly of PSI-FCPI. Importantly, the selective binding of FCPIs was identified for the first time by comparing the structures of PSI-FCPI supercomplexes and the amino acid sequences of FCPIs between the two diatom species. This approach can be extended to the LHC protein superfamily in the green and red lineages, enabling comparisons of protein structures and sequences of PSI-LHCI supercomplexes among closely related species. This, in turn, lays the foundation for elucidating the underlying mechanism of PSI-LHCI supercomplex assembly. Thus, this study will shed light on answering the evolutionary question of how LHCIs recognize their binding sites at PSI in photosynthetic organisms.

Methods

Cell growth and preparation of thylakoid membranes

The marine centric diatom T. pseudonana CCMP1335 was grown in artificial seawater supplemented with sodium metasilicate and KW21 (40) at 20°C under a photosynthetic photon flux density of 30 μmol photons m−2 s−1 provided by white LED, with bubbling of air containing 3% (v/v) CO2. The cells were harvested by centrifugation, disrupted by agitation with glass beads (41), and the thylakoid membranes were pelleted by further centrifugation. The resulting thylakoid membranes were suspended in 50 mM Mes-NaOH (pH 6.5) buffer containing 1 M betaine and 1 mM EDTA.

Purification of the PSI-FCPI supercomplex

Thylakoid membranes were solubilized with 1% (w/v) n-dodecyl-β-D-maltoside (β-DDM) at a Chl concentration of 0.5 mg mL−1 for 20 min on ice in the dark with gentle stirring. After centrifugation at 162,000 × g for 20 min at 4°C, the supernatant was loaded onto a Q-Sepharose anion-exchange column (1.6 cm inner diameter, 25 cm length) equilibrated with 20 mM Mes-NaOH (pH 6.5) buffer containing 0.2 M trehalose, 5 mM CaCl2, and 10 mM MgCl2, and 0.03% β-DDM (buffer A). The column was washed with buffer A until the eluate became colorless. Elution was performed at a flow rate of 1.0 mL min−1 using a linear gradient of buffer A and buffer B (buffer A plus 500 mM NaCl) with the following time and gradient: 0–600 min, 0–60% buffer B; 600–800 min, 60–100% buffer B; 800–900 min, 100% buffer B. The PSI-FCPI-enriched fraction was eluted at 194–247 mM NaCl, then collected and subsequently loaded onto a linear gradient containing 10–40% (w/v) trehalose in 20 mM Mes-NaOH (pH 6.5) buffer containing 5 mM CaCl2, 10 mM MgCl2, 100 mM NaCl, and 0.03% β-DDM. After centrifugation at 154,000 × g for 18 h at 4°C (P40ST rotor; Hitachi), a green fraction (Fig. S1A) was collected and concentrated using a 150 kDa cut-off filter (Apollo; Orbital Biosciences) at 4,000 × g. The concentrated samples were stored in liquid nitrogen until use.

Biochemical and spectroscopic analyses of the PSI-FCPI supercomplex

The polypeptide bands of PSI-FCPI were analyzed by SDS- PAGE with 16% (w/v) acrylamide and 7.5 M urea, following the method of Ikeuchi and Inoue (42) (Fig. S1B). The PSI-FCPI supercomplexes (4 µg of Chl) were solubilized in 3% lithium lauryl sulfate and 75 mM dithiothreitol for 10 min at 60°C, and then loaded onto the gel. A standard molecular weight marker (SP-0110; APRO Science) was used. The absorption spectrum of PSI-FCPI was measured at room temperature using a UV-Vis spectrophotometer (UV-2450; Shimadzu) (Fig. S1C), and the fluorescence emission spectrum of PSI-FCPI was measured at 77 K upon excitation at 430 nm using a spectrofluorometer (RF-5300PC; Shimadzu) (Fig. S1D). The pigment composition of PSI-FCPI was analyzed by HPLC following the method of Nagao et al. (43), and the elution profile was monitored at 440 nm (Fig. S1E).

Cryo-EM data collection

A 3-μL aliquot of the T. pseudonana PSI-FCPI supercomplex (3.0 mg of Chl mL−1) in 20 mM MES-NaOH (pH 6.5) buffer containing 0.5 M betaine, 5 mM CaCl2, 10 mM MgCl2, and 0.03% β-DDM was applied to Quantifoil R1.2/1.3 Cu 300 mesh grids in the chamber of FEI Vitrobot Mark IV (Thermo Fisher Scientific). The grid was then blotted with filter paper for 4 sec at 4°C under 100% humidity and plunged into liquid ethane cooled by liquid nitrogen. The frozen grid was transferred to a CRYO ARM 300 electron microscope (JEOL) equipped with a cold-field emission gun operated at 300 kV. All image stacks were collected from 5 × 5 holes per stage adjustment to the central hole and image shifts were applied to the surrounding holes while maintaining an axial coma-free condition. The images were recorded using an in-column energy filter with a slit width of 20 eV at a nominal magnification of × 60,000 on a direct electron detector (Gatan K3, AMETEK). The nominal defocus range was −1.8 to −1.2 μm, and the physical pixel size corresponded to 0.752 Å. Each image stack was exposed at a dose rate of 21.46 e−Å−2sec−1 for 2.33 sec in CDS mode, with dose-fractionated 50 movie frames. A total of 8,950 image stacks were collected.

Cryo-EM image processing

The resultant movie frames were aligned and summed using MotionCor2 (44) to produce dose-weighted images. The contrast transfer function (CTF) estimation was performed using CTFFIND4 (45). All subsequent processes were carried out using RELION-4.0 (46). A total of 2,733,572 particles were automatically picked and subjected to reference-free 2D classification. From these, 1,132,721 particles were selected from well-defined 2D classes and further processed for 3D classification without imposing any symmetry. An initial model for the first 3D classification was generated de novo from the 2D classification. A 240-Å spherical mask was used during the 3D classification and refinement processes. As illustrated in Fig. S2C, the final PSI-FCPI structure was reconstructed from 75,667 particles. The overall resolution of the cryo-EM map was determined to be 2.30 Å, based on the gold-standard FSC curve with a cut-off value of 0.143 (Fig. S3A) (47). Local resolutions were calculated using RELION (Fig. S3C).

Model building and refinement

Two types of the cryo-EM maps were employed for the model building of the PSI-FCPI supercomplex: a postprocessed map and a denoised map generated using Topaz version 0.2.4 (48). The postprocessed map was denoised using a trained model over 100 epochs using two half-maps. Initial models of each subunit in the PSI-FCPI supercomplex were generated by ModelAngelo (38) and subsequently inspected and manually adjusted against the maps with Coot (49). Each model was built based on interpretable features from the density maps at a contour level of 2.5 σ in both the denoised and postprocessed maps. For the assignment of Chls, Chls a and c were distinguished by inspecting the density map corresponding to the phytol chain at the least level not to link the map of Chls with that of noise. All Chls c were assigned as Chl c1 due to the inability to distinguish between Chl c1 and Chl c2 at the present resolution. For the assignment of Cars, Fx and Ddx were distinguished based on the density surrounding the head groups of Cars with the above threshold. The PSI-FCPI structure was refined using phenix.real_space_refine (50) and Servalcat (51), incorporating geometric restraints for protein-cofactor coordination. The final model was validated with MolProbity (52), EMRinger (53), and Q-score (54). The statistics for all data collection and structure refinement are summarized in Table S1, S2. All structural figures were prepared using PyMOL (55), UCSF Chimera (56), and UCSF ChimeraX (57). Since the numbering of Chls, Cars, and other cofactors in this paper differs from those in the PDB data, the corresponding relationships are provided in Table S6–S8.

Phylogenetic analysis

Amino acid sequences were aligned using MAFFT L-INS-i v7.490 or MAFFT E-INS-i v7.520 (58). The alignment was trimmed using ClipKit v1.4.1 with the smart-gap mode. Phylogenetic trees were inferred using IQ-TREE 2 (59) with the model selected by ModelFinder (60). The trees were visualized using iTOL v6 (61). Ultrafast bootstrap approximation was performed with 1000 replicates (62).

Data availability

Atomic coordinates, cryo-EM maps, and raw image data for the reported structure have been deposited in the Protein Data Bank under an accession code 8XLS [https://www.rcsb.org/structure/8XLS], in the Electron Microscopy Data Bank under an accession code EMD-38457 [https://www.ebi.ac.uk/emdb/EMD-38457], and in the Electron Microscopy Public Image Archive under an accession code EMPIAR-12142 [https://doi.org/10.6019/EMPIAR-12142], respectively.

Supporting information